---
title: "Untitled"
author: "abbey yatsko"
date: "2024-06-27"
output: html_document
---
# libraries 
```{r}
library(dplyr)
library(tidyr)
```

# data load 
```{r}
load("/Users/abbeyyatsko/Desktop/FE_repo/data/cookie.RData")
load("/Users/abbeyyatsko/Desktop/FE_repo/data/cookie_d.RData")

load("/Users/abbeyyatsko/Desktop/FE_repo/data/biomass.RData")
load("/Users/abbeyyatsko/Desktop/FE_repo/data/tree_level_damage.RData")

load("/Users/abbeyyatsko/Desktop/FE_repo/data/cookie_heights_added_out.RData")

all_biomass <- read.csv("/Users/abbeyyatsko/Desktop/FE_repo/data/Iron_Range_Stem_Damage_master_June_06_23.xlsx - all.csv")

alltermites <- read.csv("/Users/abbeyyatsko/Desktop/FE_repo/data/termites_paul - all_ID_new.csv")

load("/Users/abbeyyatsko/Desktop/repos/ironrange/R/data_out/field_wd_species.RData")
```

# data format - comparison of microbe/termite damage at cookie/tree levels
```{r}
# cookie df formatting 
cookie$holrot <- ifelse(cookie$hollowed == "y" & cookie$rotted == "y", "y", "n")
cookie$treepart <- ifelse(cookie$position %in% c("S", "C", "A"), "main stem", "canopy")
cookie <- cookie %>% relocate(treepart, .after=position)

# create 4 categories: termite_only, microbe_only, both, no_damage
cookie$termite_only <- ifelse(cookie$hollowed == "y" & cookie$rotted == "n", "y", "n")
cookie$microbe_only <- ifelse(cookie$hollowed == "n" & cookie$rotted == "y", "y", "n")
cookie$both <- cookie$holrot
cookie$no_damage <- ifelse(cookie$hollowed == "n" & cookie$rotted == "n", "y", "n")

# df for chi square test
cookie_chi <- cookie[27:30]
cookie_chi_long <- gather(cookie_chi, key = "damage", value = "value")
cookie_chi_long <- cookie_chi_long %>%
  filter(value != 'n')
cookie_chi_long$scale <- "cookie"
# save(cookie_chi_long, file="/Users/abbeyyatsko/Desktop/FE_repo/R/data_out/cookie_chi_long.RData")

# tree df formatting
trees <- trees[c("tree_id", "ply_model", "species", "dbh_observed_cm_tidy")]
trees <- (left_join(trees, dam_scaled, by = c("ply_model", "species")))
trees$mu_tot_scaled[is.na(trees$mu_tot_scaled)] <- 0
trees$mu_rot_scaled[is.na(trees$mu_rot_scaled)] <- 0
trees$mu_hol_scaled[is.na(trees$mu_hol_scaled)] <- 0

# classify damage/no damage by tree average
trees$damaged <- ifelse(trees$mu_tot_scaled > 0, "Y",  "N")
trees$rotted <- ifelse(trees$mu_rot_scaled > 0, "Y",  "N")
trees$hollowed <- ifelse(trees$mu_hol_scaled > 0, "Y",  "N")
trees$holrot <- ifelse(trees$mu_hol_scaled > 0 & trees$mu_rot_scaled > 0, "Y", "N")

# tree level: create 4 categories: termite_only, microbe_only, both, no_damage
trees$termite_only <- ifelse(trees$hollowed == "Y" & trees$rotted == "N", "y", "n")
trees$microbe_only <- ifelse(trees$hollowed == "N" & trees$rotted == "Y", "y", "n")
trees$both <- trees$holrot
trees <- trees %>% mutate(both = tolower(both))
trees$no_damage <- ifelse(trees$hollowed == "N" & trees$rotted == "N", "y", "n")

# df for chi square test 
trees_chi <- trees[13:16]
trees_chi_long <- gather(trees_chi, key = "damage", value = "value")
trees_chi_long <- trees_chi_long %>%
  filter(value != 'n')
trees_chi_long$scale <- "tree"
# save(trees_chi_long, file="/Users/abbeyyatsko/Desktop/FE_repo/R/data_out/trees_chi_long.RData")

# merge cookie  and tree level df
merged_long <- rbind(cookie_chi_long, trees_chi_long)
freq_summary <- data.frame(table(merged_long))
freq_summary$proportion <- ifelse(freq_summary$scale == "cookie", freq_summary$Freq/200, freq_summary$Freq/63)
# save(freq_summary, file="/Users/abbeyyatsko/Desktop/FE_repo/R/data_out/freq_summary.RData")
```

# data format - microbe/termite damage at different positions in the tree 
```{r}
# convert percentage to proportion for values 0 < x < 1 
cookie_heights_added_out$prop_hollow <- cookie_heights_added_out$perc_hollow/100
cookie_heights_added_out$prop_rot <- cookie_heights_added_out$perc_rot/100

# Adding a small constant to avoid 0 and 1 values (Smithson & Verkuilen 2006)
cookie_heights_added_out$prop_hollow_tr <- 
  (cookie_heights_added_out$prop_hollow * (nrow(cookie_heights_added_out) - 1) + 0.5) / nrow(cookie_heights_added_out)
cookie_heights_added_out$prop_rot_tr <- 
  (cookie_heights_added_out$prop_rot * (nrow(cookie_heights_added_out) - 1) + 0.5) / nrow(cookie_heights_added_out)

# save(cookie_heights_added_out, file="/Users/abbeyyatsko/Desktop/FE_repo/R/data_out/cookie_heights_added_out.RData")
```

# data format - species differences in microbe/termite damage
```{r}
# for damaged trees, what is the extent of whole tree damage on behalf of different decomposers 
damaged_trees_long <- gather(dam_scaled, condition, measurement, mu_hol_scaled:mu_rot_scaled)
damaged_trees_long_rep_sp <- damaged_trees_long %>% 
  filter(species %in%  c("Corymbia sp. (clarksoniana/novoguinensis) ", "Eucalyptus tetrodonta","Lophostemon suaveolens",  "Planchonia careya"))
# save(damaged_trees_long_rep_sp, file="/Users/abbeyyatsko/Desktop/FE_repo/R/data_out/damaged_trees_long_rep_sp.RData")
```

# data format - supplementary table of plot level damaged biomass
```{r}
all_biomass <- all_biomass[, c('ply_model', 'species', 'obs_ABG_dry_kg')]
all_biomass <- all_biomass[complete.cases(all_biomass), ]

totbiomass <- sum(all_biomass$obs_ABG_dry_kg)

out <- all_biomass %>% group_by(species) %>% 
  summarize(total_biomass = sum(obs_ABG_dry_kg),
            perc_biomass = sum(obs_ABG_dry_kg)/totbiomass*100,
            n = n())

# divide total biomass by 2.1ha to get per hectare values
```

# data format - termites sampled 
```{r}
tree_termites <- alltermites[alltermites$niche == "trees",]
table(tree_termites$species_organized)

dw_termites <- alltermites[alltermites$niche == "deadwood",]
table(dw_termites$species_organized)
```

# data format - tree-level factors
```{r}
# merge in tree size 
tree_size <- trees[c("ply_model", "dbh_observed_cm_tidy")]
damaged_trees <- left_join(dam_scaled, tree_size, by = "ply_model")

# merge in wood density
damaged_trees <- left_join(damaged_trees, field_wd_species, by = c("tree_ID" = "tree_id", "species"))

# save(damaged_trees, file="/Users/abbeyyatsko/Desktop/FE_repo/R/data_out/damaged_trees.RData")
```

# data format - supplementary table 1
```{r}
# total n each species
table(dam_scaled$species)

# n for different types of damage 
dam_scaled %>% group_by(species) %>% count(mu_hol_scaled > 0 )
dam_scaled %>% group_by(species) %>% count(mu_rot_scaled > 0 )

# species level tree level damage mean and sd 
trees_d <- trees[trees$damaged == "Y",]
trees_d_stats_sp <- dam_scaled %>% group_by(species) %>% 
  summarize(mean_mic = mean(mu_rot_scaled), 
            mean_term = mean(mu_hol_scaled), 
            sd_mic = sd(mu_rot_scaled), 
            sd_term = sd(mu_hol_scaled), 
            n = n())

# overall damage
trees_d_stats <- dam_scaled %>% 
  summarize(mean_mic = mean(mu_rot_scaled), 
            mean_term = mean(mu_hol_scaled), 
            sd_mic = sd(mu_rot_scaled), 
            sd_term = sd(mu_hol_scaled), 
            max_term = max(mu_hol_scaled), 
            max_mic = max (mu_rot_scaled),
            n = n())

# how many trees are represented in the cookie df
tree_summary <- cookie %>%
  count(tree_ID, name = "count") %>%
  filter(count > 1)
```

# data format - damage throughout tree for microbe only
```{r}
# 4 trees with microbe only damage 
mo <- trees[trees$microbe_only == "y",]
mo_id <- mo$tree_ID

# subset cookie df with heights added 
mo_cookie <- cookie_heights_added_out[cookie_heights_added_out$tree_ID %in% mo_id, ]

# save(mo_cookie, file="/Users/abbeyyatsko/Desktop/FE_repo/R/data_out/mo_cookie.RData")
```

